---
title: "data wrangling"
format: 
  html:
    embed-resources: true
---

# Clean and wrangle data
## Setup
```{r}
library(dplyr)
library(tidyr)
library(readr)
```

## Source data
### Read data
```{r}
#| label: code chunk name
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

```{r}
colnames(catch_original)
head(catch_original)
summary(catch_original)
glimpse(catch_original)
```

## Clean and reshape data

```{r}
catch_data <- catch_original %>% 
    select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
```

NB you can use select to change colnames as you select, e.g. 
```{r}
catch_data2 <- catch_original %>% 
    select(Region, Year, Total = All)
```

Note that designing data tables with `starts_with()` and `ends_with()` in mind is smart becuase it's a good way
to help cleaning because you can easily select/exclude certain types of column i.e. variable

### Chinook should not be char
```{r}
catch_clean <- catch_data %>% 
    mutate(Chinook = if_else(
        Chinook == "I",
        "1", # can't go straight to numeric because data is currently chr
        Chinook
    )) %>% 
    mutate(Chinook = as.numeric(Chinook))

catch_clean2 <- catch_data %>% 
    mutate(Chinook = if_else(
        Chinook == "I",
        1,
        as.numeric(Chinook)
    ))

which(is.na(catch_clean$Chinook))
```

### Go from wide to long
```{r}
catch_long <- catch_clean %>% 
    pivot_longer(col = -c(Region, Year),
        names_to = "species",
        values_to = "catch_thousands"
    )
```

### Long to wide
```{r}
catch_wide <- catch_long %>% 
    pivot_wider(
        names_from = species,
        values_from = catch_thousands,
        values_fill = 0 # anything missing is 0
    )
head(catch_wide)
```

### Clean long - make the catch in thousands
```{r}
catch_long <- catch_long %>% # careful overwriting here, only do if doing something super simple
    mutate(catch = catch_thousands * 1000) %>% 
    select(-catch_thousands)
head(catch_long)
```

```{r}
mean_region <- catch_long %>% 
    group_by(Region) %>% # doesn't change data, just creates a kind of internal flag identifying and storing the groups
    summarize(mean_catch = mean(catch, na.rm = TRUE), .groups = "drop") %>% 
    arrange(desc(mean_catch)) # else default is to alphabetize by grouped variable
head(mean_region)

# good practice to remove NA even if you already have! .groups drop ungroups after doing the operation
```

## Filter
```{r}
sse_catch <- catch_long %>% 
    filter(Region == "SSE")

head(sse_catch)
```

## Split, combine

CHECK THIS
```{r}
sites_df <- data.frame(site = c("HAW-101", "HAW-103", "OAH-320", "OAH-219", "MAU-039"))

sites_df <- sites_df %>% 
    separate_wider_delim(
        cols = site, 
        delim = "-",
        names = c("island", "site_number"),
        cols_remove = FALSE)

# Check that out
head(sites_df)
```


```{r}
dates_df %>% 
    unite(
        col = date,
        year, month, day,
        sep = "-",
        remove = FALSE
    )
```


# Put it all together

```{r}
library(dplyr)
library(tidyr)
library(readr)

# grab the data
catch_original <- read_csv(file = "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1",
    show_col_types = FALSE)

# Drop unwanted columns, fix "I" typo, reshape to long, transform catch
# into individual fish count, calculate mean catch by region, and sort
# in descending order
mean_region <- catch_original %>%
    select(-All, -notesRegCode) %>% 
    mutate(
        Chinook = if_else(Chinook == "I", 
            true = "1", false = Chinook),
        Chinook = as.numeric(Chinook)) %>% 
    pivot_longer(-c(Region, Year), 
        names_to = "species", 
        values_to = "catch") %>%
    mutate(catch = catch*1000) %>% 
    group_by(Region) %>% 
    summarize(mean_catch = mean(catch, na.rm = TRUE),
        .groups = "drop") %>% 
    arrange(dplyr::desc(mean_catch))



# Check the result
head(mean_region)

```


